Globally-centered autocovariances in MCMC

### 2020-07-22 Paper/main.tex
First superficial change

\usepackage{amssymb,amsthm,amsmath}
\usepackage{commath}
\usepackage{enumerate,xcolor}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{caption,setspace}
\usepackage{subcaption}
\DeclareMathOperator{\Tr}{Tr}
\newcommand{\dd}[1]{\mathrm{d}#1}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\B}{\mathcal{B}}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
% These three packages are from the American Mathematical Society and includes all of the important symbols and operations 
\usepackage{fullpage}
\usepackage{comment}
% By default, an article has some vary large margins to fit the smaller page format.  This allows us to use more standard margins.
\allowdisplaybreaks
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

@@

{\color{blue} Main things to change
\begin{itemize}
  \item I don't want to call this ``Replicated autocovariance'' or ``Replicated spectral variance''. I think ``Globally-centered autocovariance'' is more appropriate and self-explanatory. So we will have to go over the whole document and change this.

\end{itemize}

A rough outline of the organization
\begin{enumerate}
  \item Introduction, with a sneak peak at new plots. Change this example to a univariate mixture of Gaussians.

  \item Globally-centered autocovariance
    \begin{itemize}
      \item Theoretical soundness
    \end{itemize}

  \item Variance of Monte Carlo averages

    Describe MC averages and CLT
    \begin{itemize}
      \item Globally-centered Spectral variance estimators
      \begin{itemize}
        \item Theoretical results
        \item Computational gains
      \end{itemize}
      \item Explore: Charlie's initial sequence estimators
    \end{itemize}

  \item Effective sample size

  \item Examples

  \item Appendix
    \begin{itemize}
      \item Proofs
      \item ACF plots for other examples.
    \end{itemize}

\end{enumerate}
}
\section{Introduction} \label{sec:intro}

Markov chain Monte Carlo (MCMC) is a sampling method used to obtain correlated realizations from a stochastic process. Due to this serial correlation in observations, a large number of MCMC samples are required to estimate features of the target distribution as compared to the independent ones. Therefore, it becomes imperative for an MCMC practitioner to have an idea about underlying dependence structure in the Markov chains. Autocovariance function (ACF) characterizes this dependence for a stationary Markov chain. A famous sample autovariance function is the empirical estimator, which unfortunately comes with its drawbacks. Firstly, it is not immune to outliers (see \cite{ma2000highly}) and secondly, it is usually non-informative in case of slowly mixing Markov chains for decently finite samples. We address the second issue in this paper by using multiple representative samples of the target distribution via parallel Markov chains.

Calculation of exact theoretical ACF has been done for simple models like autoregressive (AR), moving average (MA) (\cite{quenouille1947notes}), and autoregressive-moving-average (ARMA) (\cite{box2015time}) models. However, for most of the complex processes, we rely on sample ACF to understand the correlation in the chain at a particular lag. For a slowly mixing Markov chain or a multi-modal target distribution, Markov chains explore the sample space only partially for even high sample size. In such situations, single chain empirical ACF will underestimate the autocovariance giving misleading inferences. With convenient multi-core computation resources, it is possible to get multiple Markov chains for a target distribution.  We propose a new autocovariance estimator called replicated autocovariance function (R-ACF) with a simple fix in centering a chain about the overall mean. An important application of ACF is in calculating the asymptotic variance in the CLT of a Markov chain process.

Suppose we have $m$ replications of Markov chains with $n$ samples in each chain. Let $\{X_{st}\}_{t \geq 1}$ denote the $s^{th}$ Harris ergodic (aperiodic, F-irreducible, and positive Harris recurrent) Markov chain having invariant distribution $F$. We will use the following Monte Carlo estimator in order to estimate $\mu$ for the $s^{th}$ chain
%
\[
\hat{\mu}_s = \dfrac{1}{n}\sum_{i = 1}^{n} g(X_{si}) \xrightarrow{a.s.} \mu \textrm{ as } n \to \infty
\]

Throughout this paper, we will deal with the function $g(X) = X$. We denote the sample mean of $s^{th}$ Markov chain by $\bar{X}_s$ and the overall mean by $\bar{\bar{X}} = \sum_{s = 1}^{m}\bar{X}_s/m$. We are interested in Monte Carlo error, i.e. $\bar{\bar{X}} - \mu$ to assess the reliability of simulations. The sampling distribution for Monte Carlo error of $s^{th}$ chain is available via Markov chain CLT (\cite{jones2004markov}) if there exists a positive definite matrix $\Sigma$ such that
%
\[
\sqrt{n}(\bar{X}_s-\mu) \xrightarrow{d} N(0,\Sigma)
\]

 With $m$ i.i.d. samples of $\bar{X}_s$, $s \in \{1,..., m\}$, the MCSE is given by $\Sigma/mn$. We are interested in consistent estimates of $\Sigma$ for two reasons. Firstly, to construct asymptotically valid confidence intervals and secondly, to determine when to stop the simulations. For this purpose, we use the class of Multivariate Spectral Variance Estimators (MSVE) wherein we use R-ACF to estimate the autocovariance. We call this replicated spectral variance (RSV) estimator (see \cite{arg:and:2006} for replicated batch means).

MCMC ensures asymptotic convergence, however this is not practically very useful where the number of samples is finite. There are many criteria for terminating simulation like fixed-time rule, sequencial fixed-volume rule (see \cite{glynn1992asymptotic}), and using Gelman Rubin diagnostics (\cite{gelman1992inference}). The relative fixed-volume rule has been thoroughly discussed for MCMC by \cite{flegal2015relative} and \cite{gong2016practical}. We will use method of effective sample size (ESS) defined by \cite{vats2019multivariate} for terminating the simulations. The quality of estimation of $\Sigma$ is critical in determining that the simulations are not stopped prematurely.

Markov chains are serially correlated and therefore $\Var_F(X)$ is non-informative about the correlatons in Markov chain. It is important to construct an estimator for $\Sigma$ which is very close to underlying reality. There are many techniques for estimating $\Sigma$ like batch means (BM), overlapping bath means (OBM), and spectral variance estimator (SVE). Due to their highly accurate results, we will only focus on SV estimates in this paper. Strong consistency for MSVE has been addressed by \cite{vats:fleg:jon:2018}. Bias and variance calculations for a variant of SV estimator with known mean is done by \cite{hannan2009multiple}. In this paper, we will address the asymptotic properties (basically strong consistency), bias and variance calculations for RSV estimator.

The most common practice would be to combine the SVE from $m$ chains by averaging over them. We call this technique average spectral variance (ASV) estimator. We will address the issue of slowly mixing Markov chains and show that RSV performs better than ASV. Whereas on one side RSV gives better estimates than ASV for a poor Markov chain, it is as good as ASV for a good Markov chain. We will exemplify this property of RSV for a bi-modal target distribution in subsection \ref{ex:boomerang}. This appeals intuitively because for a fast mixing Markov chain, all the chains will give almost same estimates for mean.

In the next section we introduce the new sample autocovariance estimator and calculate its bias for finite samples. We observe that the bias term for R-ACF is similar to the bias of empirical ACF except for a few small order end-effect terms that vanish as $n \to \infty$. In section \ref{sec:variance_est}, we introduce three estimators for $\Sigma$ and study the properties of RSV estimator through proof of strong consistency, bias, and variance calculations. We will elaborate on the estimator for ESS using RSV in section \ref{sec:ess}. Section \ref{sec:examples} gives a simulation study on three different examples for the proposed estimators. There are only a handful of stochastic processes where the true autocovariance and $\Sigma$ are known. To compare the estimators when the truth is known, we use a slowly mixing vector autoregressive process of order 1 (VAR(1)). We successfully show in section \ref{ex:var} that our estimators yield better results as compared to the classical sample ACF and SVE. A more promising advantage of our estimators is observed when the target distribution is multi-modal. For this purpose we use a bivariate bi-modal probability distribution introduced by \cite{gelman1991note} in section \ref{ex:boomerang}. All the important proofs are given in the Appendix.

\begin{example}[Motivating example]
  gj
\begin{figure}
\begin{subfigure}{\textwidth}
  \centering
  \includegraphics[width=.8\linewidth]{plots/acf,n=500.pdf}
  \caption{$n = 500$}
  \label{subfig:acf-500}
\end{subfigure}\\
\begin{subfigure}{\textwidth}
  \centering
  \includegraphics[width=.8\linewidth]{plots/acf,n=10000.pdf} 
  \caption{$n = 10^4$}
  \label{subfig:acf-5e4}
\end{subfigure}
\caption{Autocorrelation plot for ACF and R-ACF for first chain out of two parallel Markov chains. First column corresponds to ACF calculated using arithmetic mean of first Markov chain and the second column corresponds to the one calculated using global mean of two Markov chains.(a): $n = 500$, not converged yet; (b): $n =  10^4$, converged}
\label{fig:var_acf_ccf}
\end{figure}

\end{example}
\section{Globally-centered autocovariance} \label{sec:R-ACF}

For a stationary stochastic process $\{X_t\}_{t \geq 1}$, let $\Gamma(k)$ denote the lag -k autocovariance. In time series literature, we generally encounter the following estimator for autocovariance. 
%
\[
\hat{\Gamma}(k) = \dfrac{1}{n}\sum_{t=1}^{n-\abs{k}}(X_t - \bar{X})(X_{(t+\abs{k})} - \bar{X})^T
\]

\cite{priestley1981spectral} shows that this estimator is biased with the main bias term equal to $\abs{k}\Gamma(k)/n$, ignoring the small order terms arising due to estimation of $\mu$. An unbiased estimator with a divisor of $n - \abs{k}$ instead of $n$ exists, say $\Gamma^*(k)$. Nevertheless, the biased estimator is preferred for its lower mean square error (\cite{priestley1981spectral}). For a small $k$ relative to $n$, the two estimators are almost the same. However, for larger relative $k$ the variance of $\hat{\Gamma}(k)$ compensates for larger bias rendering a smaller mean square error than $\Gamma^*(k)$.

A commonly faced problem with the MCMC algorithms is the slow mixing property. In case of multi modal target distributions, it has been observed that even decently mixing Markov chains fail to explore the sample space for the first few thousand observations. This leads to underestimation of autocovariance for a single Markov chain. We propose a simple and neat fix for estimating ACF while dealing with slow convergence. Suppose we have $m$ replications of Markov chains for same target distribution, the overall mean $\bar{\bar{X}} = \sum_{i = 1}^{m}\bar{X}_i/m$ provides a better estimate for the expectation value when the number of simulation per chain are not sufficient for the convergence to kick in. The replicated autocovariance estimator or $s^{th}$ Markov chain is given by:
%
\[
\hat{\Gamma}_{RAC,s}(k) = \dfrac{1}{n}\sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T
\]

If the starting points of these parallel Markov chains are sufficiently dispersed, R-ACF provides more realistic estimation of lag covariances. It is a known problem that getting true autocovariance in closed form for non-simple problems is difficult. In Figure \ref{fig:var_acf_ccf}, we will illustrate the comparison between ACF and R-ACF for a slowly mixing VAR-1 example, where the truth is known. We wish to make two points here: firstly, for small sample size, we observe that ACF gives extremely misleading estimates of autocovariance whereas R-ACF is very close to reality. Figure \ref{subfig:acf-500} depicts the scenario where the convergence is not achieved yet. Secondly, for a large sample size , the estimates from R-ACF as well as ACF are equivalent. Figure \ref{subfig:acf-5e4} depicts the scenario when convergence is achieved. This shows that R-ACF is at least as good as ACF; hence safe to use in any setting.

The bias for R-ACF is similar to the bias of sample autocovariance estimator in case of single chain except for a few small order terms that shall be later addressed in theorem \ref{th:RAC_expec}. Before getting to the bias results for $\hat{\Gamma}_{RAV}(k)$ for any lag k, we introduce an additional notation. For $q \geq 1$, we define
\[
\Phi^{(q)} = \sum\limits_{-\infty}^{\infty}\abs{k}^q \|\Gamma(k)\|
\]
We denote $\Phi^{(1)} \textrm{ by } \Phi$. 
\begin{theorem} \label{th:RAC_expec}
   \[
   \mathbb{E}[\hat{\Gamma}_{RAC}(k)] = \left(1- \dfrac{|k|}{n}\right)\left(\Gamma(k) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2}\right)  + \dfrac{2(m-1)}{m}\dfrac{\abs{k}}{n^2} \left(\Sigma - \sum_{h=0}^{n-1}\Gamma(h)\right) + o(n^{-2})
   \]
\end{theorem}

@@

As a consequence, if $\Var(\bar{X})$ is finite, then $\Var(\bar{X}) \to 0 \textrm{ as } n \to \infty$. We can break $\hat{\Gamma}_{RAC}$ into four parts for all $k \geq 1$ as:
 \[
 \hat{\Gamma}_{RAC} =  \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Gamma}_s(k) - \dfrac{1}{mn}\sum_{s=1}^{m}\sum_{t=1}^{|k|}A_{st}^T - \dfrac{1}{mn}\sum_{s=1}^{m}\sum_{t=n-|k|+1}^{n}A_{st} + \dfrac{n-|k|}{mn}\sum_{s=1}^{m}(\bar{X}_s - \bar{\bar{X}})(\bar{X}_s - \bar{\bar{X}})^T\,,
 \]
%
where $A_{st} = (X_{st}-\bar{X}_s)(\bar{X}_s - \bar{\bar{X}})^T$\\
The expectation value of the first term is given by equation \ref{eq:priestly} and \ref{eq:song} while the other three terms contribute $o(1)$ terms. The complete proof evaluating the expectation values of last three terms can be found in Appendix subsection \ref{appendix:bias}.
\end{proof}

\begin{remark} \label{rmrk:exp_racf_minus_acf}
Using theorem \ref{th:RAC_expec} and equation \ref{eq:priestly},
\[
\mathbb{E}[\hat{\Gamma}_{RAC}(k) - \hat{\Gamma}(k)] = \dfrac{\abs{k}}{n}\left(1 - \dfrac{1}{m}\right)\left(\dfrac{\Sigma}{n} + \dfrac{\Phi}{n^2}\right) + \dfrac{2(m-1)}{m}\dfrac{\abs{k}}{n^2} \left(\Sigma - \sum_{h=0}^{n-1}\Gamma(h)\right) + o(n^{-2})
\]
We know that the sample autocovariance $\hat{\Gamma}(k)$ underestimates the truth due to a negative bias of $\abs{k}\Gamma(k)/n$ for finite $n$. However, in practice, it is preferred to overestimate the autocovariance for finite $n$ than to underestimate it in order to have a more realistic idea about the correlation in a Markov chain at a particular lag.  Note that the RHS in the above expression is positive. As a consequence, R-ACF will counter this negative bias and give more reliable estimates for ACF.

@@

\begin{corollary} \label{cor:lag0_expectation}
\[
\mathbb{E}[\hat{\Gamma}_{RAC}(0)] = \Gamma(0) + \dfrac{\Sigma}{mn} + \dfrac{\Phi}{mn^2} + o(n^{-2})
\]
\end{corollary}

@@

Let $\{X_t\}_{t \geq 1}$ be a stochastic process. Denote lag-k autocovariance matrix by $\Gamma(k)$, which is given by,

\[
\Gamma(k) = \Gamma(-k) = \E[(X_t - \E[X])(X_{t+k} - \E[X])^T]
\]

Recall that $\hat{\Gamma}(k)$ is the empirical autocovariance function given by

\[
\hat{\Gamma}(k) = \dfrac{1}{n}\sum_{t=1}^{n-\abs{k}}(X_t - \bar{X})(X_{t + \abs{k}} - \bar{X})^T
\]

@@

\[
\hat{\Sigma}_{SV} = \sum_{k=-b_n+1}^{b_n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Gamma}_n(k)
\]

where $w(.)$ is the lag window and $b_n$ is the truncation point.

 Consider $m$ Markov chains where $\bar{X}_s$, $\hat{\Gamma}_s(k)$ and $\hat{\Sigma}_{SV, s}$ are the sample mean, sample lag-k autocovariance and spectral variance estimators respectively for $s^{th}$ Markov chain, $s\in {1,...,m}$. We will consider two estimators for spectral variance calculations using multiple Markov chains.

\subsection{Average Spectral Variance } \label{sec:asv}

Recall that for $s^{th}$ Markov chain, the spectral variance estimator is given by,

>>>>

\usepackage{amssymb,amsthm,amsmath}
\usepackage{commath}
\usepackage{enumerate,xcolor}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{caption,setspace}
\usepackage{subcaption}
\DeclareMathOperator{\Tr}{Tr}
\newcommand{\dd}[1]{\mathrm{d}#1}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\B}{\mathcal{B}}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
% These three packages are from the American Mathematical Society and includes all of the important symbols and operations 
\usepackage{fullpage}
\usepackage{comment}
\usepackage[linesnumbered, ruled]{algorithm2e}
% By default, an article has some vary large margins to fit the smaller page format.  This allows us to use more standard margins.
\allowdisplaybreaks
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

@@

% {\color{blue} Main things to change
% \begin{itemize}
%   \item I don't want to call this ``Replicated autocovariance'' or ``Replicated spectral variance''. I think ``Globally-centered autocovariance'' is more appropriate and self-explanatory. So we will have to go over the whole document and change this.

%   \item The contents are a bit messy right now, and I am having trouble organizing my thoughts.

% \end{itemize}

% A rough outline of the organization
% \begin{enumerate}
%   \item Introduction, with a sneak peak at new plots. Change this example to a univariate mixture of Gaussians.:

%     Here, we will just explain the problem with the old ACF and what happens by using the new ACF. We will show a plot to illustrate the impact. Then we will summarize the three places this will have an impact: (1) ACF plots (2) estimators of the asymptotic variance-covariance matrix (3) estimating effective sample size.

%   \item Globally-centered autocovariance
%     \begin{itemize}
%       \item Theoretical soundness
%     \end{itemize}

%   \item Variance of Monte Carlo averages

%     Describe MC averages and CLT here and not before
%     \begin{itemize}
%       \item Globally-centered Spectral variance estimators
%       \begin{itemize}
%         \item Theoretical results
%         \item Fast implementation
%       \end{itemize}
%     \end{itemize}

%   \item Effective sample size

%   \item Examples

%   \item Discussion
%     \begin{itemize}
%       \item All plots have been made using the package:(name). The package can be found on GitHub.
%       \item One thing to discuss is that the new autocovariances can be used in the initial sequence estimators as well.
%     \end{itemize}

%   \item Appendix
%     \begin{itemize}
%       \item Proofs
%       \item ACF plots for other examples.
%     \end{itemize}

% \end{enumerate}
% }
% \newpage

\section{Introduction} \label{sec:intro}

Markov chain Monte Carlo (MCMC) is a sampling method used to obtain correlated realizations from a stochastic process. Due to this serial correlation in observations, a large number of MCMC samples are required to estimate features of the target distribution as compared to the independent ones. Therefore, it becomes imperative for an MCMC pG-titioner to have an idea about underlying dependence structure in the Markov chains. Autocovariance function (ACF) chaG-terizes this dependence for a stationary Markov chain. Suppose $\{X_t; t \in \mathbb{Z}\}$ is a stochastic process with mean $\mu$ and follows second-order stationarity i.e. (1) $\mathbb{E}[X_t] = \mu$ is constant for all $t$, (2) $\mathbb{E}[X_t^2]$ is constant for all $t$, and (3) $\mathbb{E}[X_t X_{t+k}]$ is a function of $k$ only $\forall k,t \in \mathbb{Z}$. Then, the autocovariance function at lag-$k$ is defined as

\[
    \Gamma(k) = Cov(X_t, X_{t+k}) = \mathbb{E}[(X_t - \mu)(X_{t+k} - \mu)^T].
\]

A popular sample autocovariance function is the empirical estimator. On a finite sample $X = (X_1, X_2, ..., X_n)^T$, the autocovariance at lag-$k$ is estimated as
%
\begin{equation} \label{eq:empirical_acf}
    \hat{\Gamma}(k) = \dfrac{1}{n}\sum_{t=1}^{n-\abs{k}}(X_t - \bar{X})(X_{(t+\abs{k})} - \bar{X})^T
\end{equation}

where $\bar{X} = \sum_{t=1}^nX_t/n$. Calculation of exact ACF has been done for simple models like autoregressive (AR), moving average (MA) (\cite{quenouille1947notes}), and autoregressive-moving-average (ARMA) (\cite{box2015time}) models. However, for most of the complex sampling algorithms, we rely on sample estimators for ACF. Unfortunately, the estimator given in equation \ref{eq:empirical_acf} has been observed to have a some drawbacks. Firstly, it is not immune to outliers (see \cite{ma2000highly}) and secondly, it is usually non-informative in case of slowly mixing Markov chains for decently finite samples. We address the second issue in this paper by using multiple representative samples of the target distribution via parallel Markov chains. With convenient multi-core computation resources, it is possible to get multiple Markov chains for a target distribution. Our goal is to combine the information from all the Markov chains while calculating the autocovariances\\

 A commonly faced problem with the MCMC models is the slow mixing property. In case of multi modal target distributions, it has been observed that the Markov chains tend to get \textit{stuck} in a particular mode for a long time. As a consequence, the chain fails to explore the sample space. In such situations, single chain empirical ACF will underestimate the autocovariance giving misleading inferences.  We propose an autocovariance estimator called globally-centered autocovariance function (G-ACF) with a simple fix in centering which has been observed to show drastically better results in many cases. We will further study the impact of G-ACF in calculation of two important applications of ACF (1) asymptotic variance of Markov chain average and (2) effective sample size (ESS).\\

 MCMC ensures asymptotic convergence, however this is not practically very useful where the number of samples is finite. There are many criteria for terminating simulation like fixed-time rule, sequencial fixed-volume rule (see \cite{glynn1992asymptotic}), and using Gelman Rubin diagnostics (\cite{gelman1992inference}). The relative fixed-volume rule has been thoroughly discussed for MCMC by \cite{flegal2015relative} and \cite{gong2016practical}. We will use method of effective sample size (ESS) by \cite{vats2019multivariate} for terminating the simulations. The quality of estimation of $\Sigma$ is critical in estimating their ESS. We will successfully show that the G-SVE estimates the truth better and prevents the premature stopping of Markov chain.\\

In the next section we introduce the new sample autocovariance estimator and calculate its bias for finite samples. We observe that the bias term for G-ACF is similar to the bias of empirical ACF except for a few small order end-effect terms that vanish as $n \to \infty$. In section \ref{sec:variance_est}, we introduce the globally-centered SVE and give a proof of strong consistency, bias, and variance calculations. We will elaborate on the estimator for ESS using G-SVE in section \ref{sec:ess}. Section \ref{sec:examples} gives an experimental study on three different examples for the proposed estimators. All the important proofs are given in the Appendix.

\begin{example}[Demonstrative example]

We demonstrate the striking difference in the ACF plots through mixture of Gaussian target density using a random-walk Metropolis-Hastings sampler. The target density has two well separated and non-equal modes. Two Markov chains are started close to each mode. It can be seen in Figure \ref{fig:gaussian-trace} that for the first $10^4$ samples, the chains do not communicate with the other mode. Whereas for a sample size of $10^5$, the chains have jumped modes and therefore, a single chain can give reliable information about the distribution. We wish to make two points here: firstly, for small sample size, we observe that ACF gives extremely misleading estimates of autocovariance whereas G-ACF is very close to reality (Figure \ref{subfig:gauss-acf-1e4}). Whereas, sharing the information about Monte Carlo average across chains leads to better estimation of autocovariances. Secondly, for a large sample size, the estimates from G-ACF as well as ACF are equivalent. Figure \ref{subfig:gauss-trace_1e5} depicts the scenario when convergence is achieved. This shows that G-ACF is at least as good as ACF; hence safe to use in any setting.

\begin{figure}[htbp]
    \begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=\linewidth]{plots/gaussian-Targettrace_n1e4.pdf}
   \caption{$n = 10^4$}
   \label{subfig:gauss-trace_1e4}
 \end{subfigure}
 \begin{subfigure}{0.5\textwidth}
   \centering
   \includegraphics[width=\linewidth]{plots/gaussian-Targettrace_n1e5.pdf} 
   \caption{$n = 10^5$}
   \label{subfig:gauss-trace_1e5}
 \end{subfigure}
    \caption{Target density and Density plot for two chains starting at different values for a Gaussian mixture model with mixture probability $= 0.7, \; \mu_1 = -5, \; \mu_2 = 5, \; \sigma_1 = 1, \textrm{ and } \sigma_2 = 0.5$. (a) $n = 10^4$; chains don't interact. (b) $n = 10^5$; chains interact.}
    \label{fig:gaussian-trace}
\end{figure}

\begin{figure}[htbp]
\begin{subfigure}{\textwidth}
   \centering
   \includegraphics[width=.8\linewidth]{plots/gaussian-acf_n1e4.pdf}
   \caption{$n = 10^4$}
   \label{subfig:gauss-acf-1e4}
 \end{subfigure}\\
 \begin{subfigure}{\textwidth}
   \centering
   \includegraphics[width=.8\linewidth]{plots/gaussian-acf_n1e5.pdf} 
   \caption{$n = 10^5$}
   \label{subfig:gauss-acf-1e5}
 \end{subfigure}
 \caption{Autocorrelation plot for locally-centered ACF (left) and globally centered ACF (right) averaged over all chains. (a) $n = 10^4$, not converged yet; (b) $n =  10^5$, converged}
 \label{fig:gauss:acf}
\end{figure}
\end{example}

\section{Globally-centered autocovariance} \label{sec:G-ACF}

 Suppose we have $m$ replications of Markov chains for same target distribution and $\{X_{st}; \; t \in \mathbb{Z}\}$ denote the $s^{th}$ Markov chain. Let the sample mean of $s^{th}$ Markov chain be denoted by $\bar{X}_s$ and the global mean by $\bar{\bar{X}} = \sum_{s = 1}^{m}\bar{X}_s/m$. The global mean provides a better estimate for the expectation value when the number of simulation per chain are not sufficient for the convergence to kick in. The globally-centered autocovariance function (G-ACF) for the $s^{th}$ Markov chain is given by:
%
\[
\hat{\Gamma}_{G}(k) = \dfrac{1}{mn} \sum_{s=1}^{m}\sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T
\]

If the starting points of these parallel Markov chains are sufficiently dispersed, G-ACF provides more realistic estimation of lag covariances. \cite{priestley1981spectral} shows that the empirical estimator in Equation \ref{eq:empirical_acf} is biased with the main bias term equal to $\abs{k}\Gamma(k)/n$, ignoring the small order terms arising due to estimation of $\mu$. An unbiased estimator with a divisor of $n - \abs{k}$ instead of $n$ exists, say $\Gamma^*(k)$. Nevertheless, the biased estimator is preferred for its lower mean square error (\cite{priestley1981spectral}). For a small $k$ relative to $n$, the two estimators are almost the same. However, for larger relative $k$ the variance of $\hat{\Gamma}(k)$ compensates for larger bias rendering a smaller mean square error than $\Gamma^*(k)$.

\subsection{Theoretical results} % (fold)
\label{sub:theoretical_results}

% subsection theoretical_results (end)

The bias for G-ACF is similar to the bias of sample autocovariance estimator in case of single chain except for a few small order terms that shall be later addressed in theorem \ref{th:G-_expec}. Before getting to the bias results for $\hat{\Gamma}_{RAV}(k)$ for any lag $k$, we introduce an additional notation. For $q \geq 1$, we define
\[
\Phi^{(q)} = \sum\limits_{-\infty}^{\infty}\abs{k}^q \|\Gamma(k)\|
\]
We denote $\Phi^{(1)} \textrm{ by } \Phi$. 
\begin{theorem} \label{th:G-_expec}
   \[
   \mathbb{E}[\hat{\Gamma}_{G}(k)] = \left(1- \dfrac{|k|}{n}\right)\left(\Gamma(k) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2}\right)  + \dfrac{2(m-1)}{m}\dfrac{\abs{k}}{n^2} \left(\Sigma - \sum_{h=0}^{n-1}\Gamma(h)\right) + o(n^{-2})
   \]
\end{theorem}

@@

As a consequence, if $\Var(\bar{X})$ is finite, then $\Var(\bar{X}) \to 0 \textrm{ as } n \to \infty$. We can break $\hat{\Gamma}_{G-}$ into four parts for all $k \geq 1$ as:
 \[
 \hat{\Gamma}_{G} =  \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Gamma}_s(k) - \dfrac{1}{mn}\sum_{s=1}^{m}\sum_{t=1}^{|k|}A_{st}^T - \dfrac{1}{mn}\sum_{s=1}^{m}\sum_{t=n-|k|+1}^{n}A_{st} + \dfrac{n-|k|}{mn}\sum_{s=1}^{m}(\bar{X}_s - \bar{\bar{X}})(\bar{X}_s - \bar{\bar{X}})^T\,,
 \]
%
where $A_{st} = (X_{st}-\bar{X}_s)(\bar{X}_s - \bar{\bar{X}})^T$\\
The expectation value of the first term is given by equation \ref{eq:priestly} and \ref{eq:song} while the other three terms contribute $o(1)$ terms. The complete proof evaluating the expectation values of last three terms can be found in Appendix subsection \ref{appendix:bias}.
\end{proof}

\begin{remark} \label{rmrk:exp_G-f_minus_acf}
Using theorem \ref{th:G_expec} and equation \ref{eq:priestly},
\[
\mathbb{E}[\hat{\Gamma}_{G}(k) - \hat{\Gamma}(k)] = \dfrac{\abs{k}}{n}\left(1 - \dfrac{1}{m}\right)\left(\dfrac{\Sigma}{n} + \dfrac{\Phi}{n^2}\right) + \dfrac{2(m-1)}{m}\dfrac{\abs{k}}{n^2} \left(\Sigma - \sum_{h=0}^{n-1}\Gamma(h)\right) + o(n^{-2})
\]
We know that the sample autocovariance $\hat{\Gamma}(k)$ underestimates the truth due to a negative bias of $\abs{k}\Gamma(k)/n$ for finite $n$. However, in practice, it is preferred to overestimate the autocovariance for finite $n$ than to underestimate it in order to have a more realistic idea about the correlation in a Markov chain at a particular lag.  Note that the RHS in the above expression is positive. As a consequence, G-ACF will counter this negative bias and give more reliable estimates for ACF.

@@

\begin{corollary} \label{cor:lag0_expectation}
\[
\mathbb{E}[\hat{\Gamma}_{G}(0)] = \Gamma(0) + \dfrac{\Sigma}{mn} + \dfrac{\Phi}{mn^2} + o(n^{-2})
\]
\end{corollary}

@@

{\color{blue}Dootika: this section here needs to be much tighter in its presentation. We will need to remember that the paper is no longer about the SV estimators but about autocorrelations. So will have to present the discussion similarly. }

Let $\{X_{t}\}_{t \geq 1}$ denote the Harris ergodic (aperiodic, F-irreducible, and positive Harris recurrent) Markov chain having invariant distribution $F$. We will use the following Monte Carlo estimator in order to estimate $\mu$
%
\[
\hat{\mu}_s = \dfrac{1}{n}\sum_{i = 1}^{n} g(X_{si}) \xrightarrow{a.s.} \mu \textrm{ as } n \to \infty
\]

We are interested in Markov chain average and therefore, throughout this paper, we will deal with the function $g(X) = X$. To assess the reliability of our simulations, we are interested in Monte Carlo error, i.e. $\bar{\bar{X}} - \mu$. The sampling distribution for Monte Carlo error is available via Markov chain CLT (\cite{jones2004markov}) if there exists a positive definite matrix $\Sigma$ such that
%
\[
\sqrt{n}(\bar{X}_s-\mu) \xrightarrow{d} N(0,\Sigma)
\]

 It is important to construct an estimator for $\Sigma$ which is very close to underlying reality. There are many techniques for estimating $\Sigma$ like batch means (BM), overlapping bath means (OBM), and spectral variance estimator (SVE). Due to its nice statistical properties and heavy dependence on autocovariance, we will only focus on SVE in this paper. SVE involves a truncated and weighted sum of autocovariances. We will give theoretical as well as experimental results for the version of SVE that uses our globally centered autocovariances. We call this globally-centered SVE (G-SVE).

Given m parallel Markov chains, we have $m$ i.i.d. samples of $\bar{X}_s$, $s \in \{1,..., m\}$. The Markov Chain standard error (MCSE) is given by $\Sigma/mn$. We are interested in consistent estimates of $\Sigma$ for two reasons. Firstly, to construct asymptotically valid confidence intervals and secondly, to determine when to stop the simulations. For this purpose, we use the class of Multivariate Spectral Variance Estimators (MSVE). The most common practice would be to combine the spectral variance estimates from $m$ chains by averaging over them. We call this technique average spectral variance estimator and denote it by $\hat{\Sigma}_A$). We propose using a globally-centered spectral variance estimator (G-SVE) wherein we use G-ACF to estimate the autocovariance. Strong consistency for MSVE has been addressed by \cite{vats:fleg:jon:2018}. Bias and variance calculations for a variant of SVE with known mean is done by \cite{hannan2009multiple}. In this paper, we will address the asymptotic properties (strong consistency), bias and variance calculations for G-SVE.

\subsection{Globally-centered spectral variance estimators} % (fold)
\label{sub:globally_centered_spectral_variance_estimators}

% subsection globally_centered_spectral_variance_estimators (end)

Recall from Equation \ref{eq:empirical_acf} that $\hat{\Gamma}(k)$ is the empirical autocovariance function given by

\begin{equation}
    \hat{\Gamma}(k) = \dfrac{1}{n}\sum_{t=1}^{n-\abs{k}}(X_t - \bar{X})(X_{t + \abs{k}} - \bar{X})^T
\end{equation}

@@

\begin{equation} \label{eq:sve}
    \hat{\Sigma}_{SV} = \sum_{k=-b_n+1}^{b_n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Gamma}_n(k)
\end{equation}

where $w(.)$ is the lag window and $b_n$ is the truncation point. Consider $m$ Markov chains where $\hat{\Gamma}_s(k)$, $\hat{\Gamma}_{G,s}(k)$ and $\hat{\Sigma}_{S, s}$ are the lag-$k$ empirical ACF, G-ACF, and SVE respectively for $s^{th}$ Markov chain, $s\in {1,...,m}$. We construct the globally-centered SVE estimator $\hat{\Sigma}_{G}$ using the globally-centered ACF. Consider, 
%
\begin{align*}
    \hat{\Sigma}_{G,s} &= \sum_{k=-b_n+1}^{b_n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Gamma}_{G,s}(k)
\end{align*}
\[
\hat{\Sigma}_{G} =  \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Sigma}_{G,s}
\]

\subsubsection{Theoretical results} \label{sec:G-SVE}

The G-SVE estimator centers the data around the global mean. We are interested in proving the strong consistency, finding the bias and variance of G-SVE.
